\chapter*{Acknowledgements}
\markboth{Acknowledgements}{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}

\vspace{0.5em}

I would like to thank my advisor, Martin Odersky, for the chance to work on exciting projects and with an extraordinary team. I would also like to thank the thesis committee, Andrew Kennedy, Maurizio Cimadamore, James Larus and Viktor Kuncak for their suggestions on improving this thesis.

\vspace{0.5em}

I would like to thank the members of the LAMP laboratory, past and present, who made the work fun: Michel Schinz, Francois Garillot, Iulian Dragos, Donna Malayeri, Adriaan Moors, Toni Cunei, Gilles Dubochet, Ingo Maier, Miguel Garcia, Tiark Rompf, Lukas Rytz, Thierry Coppey, Christopher Vogt, Alex Prokopec, Hubert Plociniczak, Vojin Jovanovic, Heather Miller, Manojar Jonnalagedda, Sandro Stucki, Vera Salvisberg, Eugene Burmako, Denys Shabalin, Nada Amin, Sébastien Doeraene, Dmitry Petrashko, Samuel Grütter, Tobias Schlatter, Nicolas Stucki and Vladimir Nikolaev. For each person, I could write a paragraph or more explaining how inspiring it was to work with them. But in the interest of brevity, I will just thank everyone for the great discussions and all the fun we had both in the office and outside. I would also like to thank Danielle Chamberlain for her support on the administration side and Fabien Salvi for always having a server ready when I needed one.

\vspace{0.5em}

During the five years in the lab I had the privilege to work with some great students, supervising their semester projects: Damien Obrist, Cristian Tal\u{a}u, Nicolas Stucki, Aymeric Genêt, Cédric Bastin, Romain Beguet or their Master theses: George Kollias, Nicolas Stucki and Milos Stojanovic. They have all done great work and made important contributions to the projects they worked on. I would also like to thank our visiting professors, Ond\v{r}ej Lhot\'{a}k and Dominik Gruntz for the discussions we had and their suggestions.

\vspace{0.5em}

I would also like to thank my research collaborators from outside EFPL, namely Arvind Sujeeth, Hassan Chafi, Kevin Brown, HyoukJoong Lee (Stanford), Roland Ducournau (Uni. Montpellier), Yannis Smaragdakis and Aggelos Biboudis (Uni. of Greece) and Vincent St-Amour (Northeastern) for their suggestions and the work we've done together. I am also very grateful to Simon Peyton Jones (MSR), Matt Might (Uni. of Utah) and Matei Zaharia (Berkeley/MIT) for their write-ups and suggestions on doing research, writing papers, reviews, rebuttals and preparing presentations, which guided me throughout my PhD. Before coming to the LAMP, I had a 6-month internship in the Dependable Systems laboratory of EPFL, where I learned a great deal about sustained work, research and focusing on the right things. For this, I would like to thank George Candea, \cb{S}tefan Bucur, Cristian Zamfir and the rest of the members of DSLab. Even before DSLab, I would like to thank Vlad Rusu and Laure Gonnord (INRIA Lille) for guiding my first baby steps into the world of research.

\vspace{0.5em}

The Scala community provided the motivation and feedback to expand my work outside the research area and develop it into the miniboxing compiler plugin. I would like to especially thank Erik Osheim and Tom Sweitzer (the developers of the Spire library), Stu Hood (Twitter), Alexandru Nedelcu (Eloquentix), Philip Stutz (Uni. of Zurich) for their suggestions and feedback.

\vspace{0.5em}

Last but not least, I would like to thank my wife, Ana Lucia, for her support and love but more importantly for our son, Ioan Ureche, who was born as I was writing this thesis. My parents, Octavian and Carmen-Dorina, were there for me throughout my life, both when I was happy and when I was sad. I would also like to thank my friends, Mihai Dobrescu, \cb{S}tefan Bucur, Lyusya Yartseva, Mihai Letia, Alex Ichim the Romanian community in Lausanne for providing a pleasant environment outside the office, even when I was working under pressure.

\bigskip

\noindent\textit{Lausanne, November 16th 2015}
\hfill Vlad Ureche
